---
title: "Quantitative Genomics and Genetics 2019"
subtitle: "Computer Lab 2"
author: "Kate Harline"
date: "2/7/2019"
output: pdf_document
---

-- R Markdown   
-- Functions
-- For loops   
-- If/else statements   
-- Vector and Matrix calculation

--------------------------------------------------------------------------------------

### 0. Logistics
- Thursday Section: Our usual location is booked so we'll meet at Mann B30A
- Lab1 assignment from last week was optional, this week you'll turn in a small Rmd file by next lab section

### 1. R Markdown

#### Standard text formating
- We can \textit{italicize it using LaTeX Syntax} or *italicize it using Markdown Syntax*.
- We can also \textbf{bold things using LaTex} or **bold it using Markdown**.

#### Text Size
\huge We can make text bigger

\vspace{12pt}

\tiny We can make text smaller

\vspace{12pt}

\normalsize We can set it back to normal

\vspace{12pt}

#### New line characters and vertical spacing
Look at how new lines in Markdown translate
to new lines in the knitted document.

\vspace{12pt}

Did you catch the new line in the above chunk?

How about this one?

\vspace{12pt}
\vspace{12pt}

#### Text alignment
This text starts at the left margin.

\hspace*{1in} This text starts one inch in. 

\centerline{This line will be centered}
\vspace{12pt}


#### Lists

\begin{itemize}
  \item{This list}
  \item{is made using}
  \item{LaTeX syntax}
\end{itemize}

\vspace{12pt}

* This list
* is made using
* Markdown syntax

\vspace{12pt}

\begin{enumerate}
  \item{This is a}
  \item{numbered list}
  \item{using LaTeX}
\end{enumerate}



#### Tables

Let's create a table listing the Top 10 Dog Movies as ranked by US Weekly. 

\begin{center}
   \begin{tabular}{|l|r|} % 2 cols (left, right); vert. lines
    \hline                  % draw horizontal line
    Movie & Year Released  \\
    \hline
    Lady and the Tramp & 1955 \\
    Turner and Hooch & 1989 \\
    Beethoven & 1992 \\
    Homeward Bound: The Incredible Journey & 1993 \\
    Lassie & 1994 \\
    101 Dalmations & 1996 \\
    Air Bud & 1997 \\
    Best in Show & 2000 \\
    My Dog Skip & 2000 \\
    Marley and Me & 2008 \\
    \hline
   \end{tabular}
 \end{center}


#### Creating line breaks of different widths

(1pt thickness)

\rule{\textwidth}{1pt}

(3pt thickness)

\rule{\textwidth}{3pt}



#### LaTeX equations
- conditional probability: $Pr(X_1|X_2) = \frac{Pr(X_1 \cap X_2)}{Pr(X_2)}$
- matrix: $$A_{m,n} =
 \begin{pmatrix}
  a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
  a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
  \vdots  & \vdots  & \ddots & \vdots  \\
  a_{m,1} & a_{m,2} & \cdots & a_{m,n}
 \end{pmatrix}$$
- other matrices and delimiter syntax:

 \[
    \begin{array}{ccccc}
      17& 24&  1&  8& 15\\
      23&  5&  7& 14& 16\\
       4&  6& 13& 20& 22\\
      10& 12& 19& 21&  3\\
      11& 18& 25& 2& 9 \end{array}
\]

\[
\mbox{signum}(x)=\left\{
\begin{array}{rl}
  1 & \mbox{if $x>0$} \\
  0 & \mbox{if $x=0$} \\
  -1 & \mbox{otherwise}
    \end{array} \right.
    \]
    
- numbered equations:
\begin{equation}
f'(x=0) = \lim_{x \rightarrow x_0} \frac{f(x)- f(x_0)}{x-x_0} 
\end{equation}

- centered and unnumbered equations: \[ x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a} \]
\[\sum^{\infty}_{n=1} \frac{1}{n} \] \[\int x^3\,dx \]

- other symbols and equations: $\sum_{i}^{m}$ or $\cup$, $\in$, $\pi$



#### Code chunks
```{r}
example.vector1 <- 1:10
mean(example.vector1)
```

Code options   

- `include = FALSE` prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.   
- `echo = FALSE` prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.   

For more information on using R Markdown: http://rmarkdown.rstudio.com/lesson-1.html   

--------------------------------------------------------------------------------------

### 2. Functions

- We learned that a function is something that takes in an input and gives you an output.

- Functions require "( )", whereas data objects will have "[ ]"

- R has many built in functions for commonly used methods in statistics.

```{r, comment=NA, }
# Examples of built in functions 
example.vector1 <- c(5,2,3,7,1,1,2,9,9)
# a function that calculates the mean
mean(example.vector1) 
# a function to index specific values
which(example.vector1==3)
which(!example.vector1==3)
which(example.vector1 %in% c(1,3))
```

- We can also build custom functions. 

```{r, comment=NA}
# the syntax for declaring functions, note the {} after function()
log10_add <- function(input1,input2){   # all the inputs are specified within the ( )
  cat("This is a custom function \n")   
  cat("The inputs are = ",input1,input2,"\n")  # showing you the inputs 
  output <- log10(input1) + log10(input2)       # creating an output within the function
  cat("The output is = ",output,"\n")          # print the output
  return(output)                               # return specifies the output
}
# Now we can call our custom functions like this
log10_add(100,1000)
# Note that the variable output is not created in our workspace 
ls()
# in order to save the result of a function to a variable we have to assign it to a variable
test.output <- log10_add(100,1000)
test.output
```

**Question 1**

- Can you guess what is going to happen? 13 is returned (input = 2 + x = 11 )

```{r, comment = NA, eval = FALSE}
x <- 11
test_function <- function(y){
  output <- x + y
  return(output)
}
test_function(2) 
```

- There are many ways to set your function arguments

```{r, comment = NA, eval = FALSE}
test_function <- function(y=myInput){
  output <- x + y
  return(output)
}
myInput <- 3
test_function() #the default function argument is myInput
test_function(y=5) #we now overrtide the default to be 5
```

- Short intro to scope: what happens in the function, stays in the function
- Which means any object you want to use in the function should be set as an argument

```{r, comment = NA, eval = FALSE}
output <- 5
x <- 1
test_function <- function(y=myInput){
  output <- x + y
  cat("The output is ",output,"\n")
  return(output)
}

myOuput<-test_function(4)
print(myOuput)
print(output)
```

**Question 2**

- Can you guess what is going to happen? the function returns 15, 10 + input (5)

```{r, comment = NA, eval = FALSE}
test_function <- function(myInput=5){
  output <- 5 + myInput
  return(output)
}
test_function(m=10)
```

#### Installing and Loading packages

- We can also use functions by installing published packages if somebody else did the hard work for us.

- We can install packages that are published on CRAN by using install.packages().

```{r}
# install.packages("ggplot2")
```

- Other packages can be install following the prompts from GitHub or Bioconductor: https://github.com/brooke-watson/BRRR

- Once the installation is complete, we to load the package into your current R session in order to use it.

```{r}
library(ggplot2)
require(ggplot2)
```

- Now we can use the functions from ggplot2
- ggplot is a really great function that I promise we'll get back to soon
- For now just note there are also base R functions that form plots, but ggplot looks much better

```{r}
x <- seq(1,10)
y <- sample(1:100,10,replace=T)
ggplot(data.frame(x,y), aes(x,y)) + geom_point()
ggplot(data.frame(x,y), aes(x,y,color="red")) + geom_point() +labs(title="This is a title",subtitle="This is a subtitle")
plot(x,y)
```

--------------------------------------------------------------------------------------

### 3. for loops

- For loops are mainly used in cases where you want to do a task multiple times.
- Note the notation of R-command (parameters) { what to do} is the same as the function

```{r}
N <- 3
for ( i in 1:N ){
    cat("Processing loop number = ",i,"\n")
}
```

- You can also create a loop within a loop

```{r}
for ( outer in 1:3 ){
    cat("Processing Outer Loop #", outer, "\n")
  for ( inner in 1:2 ){
      cat("Processing |_ Inner Loop #", inner, "\n")
    }
}
```

**Question3**

- What is the final value of N ? 6

```{r, eval = FALSE}
N <- 3
for( i in 1:N){
  cat("Processing loop = ", i, "\n")
  N <- N + 1
  # cat("\tN = ", N, "\n")
}
print(N)
```


--------------------------------------------------------------------------------------

### 4. If / else statements 

- By using if and else statements you can insert condition specific executions in your script

- The code inside an if statement will only be executed when the condition is TRUE

```{r, comment = NA, eval = FALSE}
if (condition) {
  do stuff
} else {
  do stuff
}
# OR you can add more levels by using else if
if(condition){
  do stuff
} else if (condition 2){
  do stuf
} else {
  do stuff
}
```

- Here is a simple example

```{r, comment = NA}
example.vector <- seq(1,25,by= 2)
# Loop over individual elements in example.vector
for( i in example.vector){
    if( i < 10 ){
        cat(i, "is smaller than 10 \n") 
    } else if ( 10 <= i & i <= 20){
        cat(i, "is in the interval [10,20] \n")
    } else {
        cat(i, "is larger than 20 \n")
    }
}
```


--------------------------------------------------------------------------------------

### 5. Vector and Matrix calculations

- If you want to modify each element of a vector by a scalar value you can use the math operations that we have learned last week.

```{r}
example.vector1
2 * example.vector1
1 + example.vector1
example.vector1 ^2
```

- If you are interested in the dot product of two vectors you have to use a special operator

```{r}
example.vector1 %*% example.vector1
```

- The same applies for matrices 

```{r}
example.matrix1 <- matrix(c(1,1,1,2,2,2), nrow = 2, ncol = 3, byrow= TRUE)
example.matrix1
2 * example.matrix1
example.matrix1 ^ 2
example.matrix1 - 1
```

- Here is how you can do matrix calculations

```{r, comment = NA}
# t() is transposing the matrix
example.matrix1 %*% t(example.matrix1)
# Note the dimensions 2 x 3 %*% 3 x 2  = 2 x 2 
```

- Here are some useful functions that can be used in matrix calculations 

```{r, comment = NA}
# creating a diagonal matrix with the first input as values on the diagonal
diag(2,nrow = 3)
# calculating the inverse of a matrix
A <- matrix(c(2,-3,1,0.5),nrow = 2)
solve(A)
# we can check this by 
A %*% solve(A) # which results in an identity matrix 
```

--------------------------------------------------------------------------------------

### Problem

Write this in an Rmarkdown (.Rmd) file and upload it to CMS before the next Lab section. Please include your netID in the file.


Part 1: Write a function that will print every value from 2 through 10, skipping 5 and printing 0 for each multiple of 3. Demonstrate use of for and if-else statments in this function.

```{r, comment = NA}

func1 <- function(first, last, skip, mult, p){
  output <- vector()
  for (i in first:last) {
    if (i != skip){
      if(i %% mult == 0){
        output <- c(output, p)
      }
      else {
        output <- c(output, i)
      }
    }
    
  }
  return(output)
}
test1 <- func1(2, 10, 5, 3, 0)
test1
```

Part 2: Try to create a vector of the values printed above without using loops or if-else statements.

```{r, comment = NA}
v <- c(2:10)
v <- v[c(1:3,5:9)]
v <- replace(v, v%%3 == 0, 0)
v
```

