---
title: "kh694_Hw3"
author: "Kate Harline"
date: "2/16/2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{geometry}
- \usepackage{graphicx}
- \usepackage{amssymb}
- \usepackage{epstopdf}
- \usepackage{setspace}
- \usepackage{paralist}
- \usepackage{amsmath}
- \usepackage{epsfig,psfrag}
- \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\title{Quantitative Genomics and Genetics - Spring 2019 \\
BTRY 4830/6830; PBSB 5201.01}
\author{Homework 3 (version 1)}
\date{Assigned February 12; Due 11:59PM February 16}    

\maketitle
\section*{Problem 1 (Easy)}

\begin{itemize}

\item[a.]

Yes, this is a valid estimator. Any function that returns a value for a parameter is technically an estimator. However, creating an estimator that correctly represents the underlying probability distribution for the data is more difficult. $\hat{p}= 0.5$ would be correct if the coin is fair.

\item[b.] 

This would calculate the $\hat{p}$ based on the actual observations from the experiment, rather than relying on the underlying assumption that the experiment properly follows a Bernoulli distribution with ${p}= 0.5$, i.e. the coin is fair.

\end{itemize}
\section*{Problem 2 (Medium)}


```{r}
a <- 3+3
a
```
